# Prosper Loans Exploration by Gabor Sar

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(GGally)
library(colorspace)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans <- read.csv('prosperLoanData.csv')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
rotate_x <- theme(axis.text.x = element_text(angle = 90, hjust = 1))
legend_alpha_1 <- guides(colour = guide_legend(override.aes = list(alpha = 1)))
```

## Analysis

### Data Overview

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(loans)
names(loans)
```

There are 113937 loans in the dataset with 81 features.

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(loans)
```

`ListingCategory..numeric.` contains integer numbers.
Converting it to a factor could be very usefull.

I am going to map the following numbers to the following levels:

Number | Level
------ | -----
0 | Not Available
1 | Debt Consolidation
2 | Home Improvement
3 | Business
4 | Personal Loan
5 | Student Use
6 | Auto
7 | Other
8 | Baby&Adoption
9 | Boat
10 | Cosmetic Procedure
11 | Engagement Ring
12 | Green Loans
13 | Household Expenses
14 | Large Purchases
15 | Medical/Dental
16 | Motorcycle
17 | RV
18 | Taxes
19 | Vacation
20 | Wedding Loans

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans$ListingCategory = factor(loans$ListingCategory..numeric., levels = 0:20)
levels(loans$ListingCategory) <- c('Not Available',
                                   'Debt Consolidation',
                                   'Home Improvement',
                                   'Business',
                                   'Personal Loan',
                                   'Student Use',
                                   'Auto',
                                   'Other',
                                   'Baby&Adoption',
                                   'Boat',
                                   'Cosmetic Procedure',
                                   'Engagement Ring',
                                   'Green Loans',
                                   'Household Expenses',
                                   'Large Purchases',
                                   'Medical/Dental',
                                   'Motorcycle',
                                   'RV',
                                   'Taxes',
                                   'Vacation',
                                   'Wedding Loans')
```

`ListingCreationDate`, `ClosedDate` and `LoanOriginationDate` are factors.
Converting them to dates could make plotting by date easier.

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans$ListingCreationDate <- as.Date(loans$ListingCreationDate)
loans$ClosedDate <- as.Date(loans$ClosedDate)
loans$LoanOriginationDate <- as.Date(loans$LoanOriginationDate)
```

### Univariate Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(loans)
```

Most of the loans have current or completed status.

The minimum estimated return is -0.183.
The median estimated return is 0.092, and the maximum is 0.284.
This seems pretty balanced.

Most of the loans (51%) for debt consolidation.

At least 88% percent of the borrowers employed, and the median of their
employment duration is 67 months.

The median upper credit score range is 699, and the maximum is 899.
The lower and upper credit score range values are very close to each other.

The median stated monthly income is $4,667, and the maximum is $1,750,003.
This is a big difference.
92% of the incomes are verifiable.

The amount of loans are between $1,000 and $35,000, and the median is $6,500.

The median monthly loan payment is $217.7, and the maximum is $2,251.5.
This is also a big difference.

#### Loan Original Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = loans) +
  geom_histogram(binwidth = 500)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(as.factor(loans$LoanOriginalAmount), maxsum = 11)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = loans) +
  geom_histogram(binwidth = 5000)
```

Most of the loan amount values are between $1,000 and $10,000.
There are some values that are more frequent than the others.
Based on the list of the ten most frequent values, it seems like approximately
every five thousandth.
Setting the binwidth to 5000 shows a decreasing trend.

#### Loan Origination Date

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginationDate), data = loans) +
  geom_histogram(binwidth = 7) +
  scale_x_date()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginationDate), data = loans) +
  geom_histogram(binwidth = 7) +
  scale_x_date(lim = c(as.Date('2008-08-01'), as.Date('2009-10-01')))
```

There is an increasing trend in the number of loans from 2006 to late 2008 and
from late 2009 to 2014, and there is a gap between late 2008 and late 2009.
Enlarging that timeframe shows that there were almost 0 loans registered in an
approximately 10-months period.
The most likely reason for this anomaly is the subprime mortgage crisis.

#### Term

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$Term)
```

77% of the loans last 36 months (3 years), 21% last 60 months (5 years) and
1% last 12 months (1 year).

#### Monthly Loan Payment

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = MonthlyLoanPayment), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = MonthlyLoanPayment), data = loans) +
  geom_histogram(binwidth = 10) +
  xlim(0, quantile(loans$MonthlyLoanPayment, .99))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$MonthlyLoanPayment == 0)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(subset(loans, MonthlyLoanPayment == 0)$LoanStatus)
```

The majority of the monthly payments are between $0 and $500.
There are some outliers.
935 loans have a monthly payment of $0.
All of those loans are completed, defaulted or the final payment is in
progress.

#### Stated Monthly Income

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = StatedMonthlyIncome), data = loans) +
  geom_histogram()
```

The histogram of monthly incomes shows a serious outlier issue.
This supports my feelings about the big difference between the median ($4,667),
and the maximum ($1,750,003).

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = StatedMonthlyIncome), data = loans) +
  geom_histogram(binwidth = 150) +
  xlim(0, quantile(loans$StatedMonthlyIncome, .99))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$StatedMonthlyIncome == 0)
```

Limiting the values to the .99 quantile and setting the binwidth to 150 shows
a much better, positively skewed normal distribution.
The only thing that does not seem obvious is the high frequency (1394) of the 0
values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# I am going to store the subset of zero StatedMonthlyIncome loans,
# so I can investigate it deeper
loans_zero_monthly_income <- subset(loans, StatedMonthlyIncome == 0)
table(loans_zero_monthly_income$ListingCategory)
```

Listing category does not explain the zero values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans_zero_monthly_income$LoanStatus)
```

Most of these loans are completed, charged off or defaulted, but still 259 of
them are current.

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans_zero_monthly_income$IncomeVerifiable)
```

Only 64 of these loans contains a verifiable monthly income.

#### Debt to Income Ratio

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = DebtToIncomeRatio), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = DebtToIncomeRatio), data = loans) +
  xlim(0, quantile(loans$DebtToIncomeRatio, .99, na.rm = TRUE)) +
  geom_histogram(binwidth = 0.01)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(loans$DebtToIncomeRatio)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$DebtToIncomeRatio > 2.5)
```

Most of the debt income ratio values are between 0.14 and 0.32.
There are some outliers.
The maximum is 10.01, and there are 435 values greater than 2.5.

#### Credit Score Range Lower

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeLower), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$CreditScoreRangeLower)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeLower), data = loans) +
  geom_histogram(binwidth = 10) +
  xlim(c(400, 900))
```

Most of the lower credit score range values are between 600 and 800.
There are some outliers.

#### Credit Score Range Upper

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeUpper), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$CreditScoreRangeUpper)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeUpper), data = loans) +
  geom_histogram(binwidth = 10) +
  xlim(c(400, 900))
```

Most of the upper credit score range values are between 600 and 800.
There are some outliers.

This data seems almost the same as the lower credit score range values.

#### Borrower Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate), data = loans) +
  geom_histogram(binwidth = 0.005)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans %>%
  group_by(BorrowerRate) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

Most borrower rates are between 0.1 and 0.3.
The values at 0.32 and 0.35 are unexpectedly frequent.

#### Borrower APR

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = loans) +
  geom_histogram(binwidth = 0.005)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# list the most frequent BorrowerAPR values
loans %>%
  group_by(BorrowerAPR) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

Most borrower APR values are between 0.1 and 0.3.
The values at 0.32 and 0.35 are unexpectedly frequent.

It seems like there is a strong relationship between borrower rate and borrower
APR.

I would like to know if there is any explanation of those unexpectedly frequent
values.

#### Employment Status

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$EmploymentStatus)
```

89% of the borrowers employed or retired (`Employed`, `Full-time`, `Part-time`,
`Retired`, `Self-employed`), <1% `Not employed`, and there is no useful
employment information about 10% of them (`NA`, `Not available`, `Other`).

#### Employment Status Duration

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = EmploymentStatusDuration), data = loans) +
  geom_histogram(binwidth = 7)
```

Most of the employment status durations are between 0 and 200 months, and the
distribution shows a decreasing trend.

#### Estimated Return

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = EstimatedReturn), data = loans) +
  geom_histogram()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = EstimatedReturn), data = loans) +
  geom_histogram(binwidth = 0.005) +
  xlim(c(0, 0.2))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$EstimatedReturn < 0)
```

Most of the estimated returns are between 0.05 and 0.15.
There are some outliers, and 195 values are negative.

#### Listing Category

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$ListingCategory)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, ListingCategory_histogram}
ggplot(aes(x = ListingCategory), data = loans) +
  geom_histogram() +
  rotate_x
```

Most loans have a listing category of debt consolidation.
A lot of loans does not have a useful listing category (Not Available, Other).

#### Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(loans$LoanStatus)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanStatus), data = loans) +
  geom_histogram() +
  rotate_x
```

Most of the loans have a loan status of current, completed, charged off or
defaulted.

### Bivariate and Multivariate Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE}
# the following values are not in the following table:
# LoanOriginationDate
# IncomeVerifiable
# * DebtToIncomeRatio
# * CreditScoreRangeLower
# * CreditScoreRangeUpper
# EmploymentStatus
# EmploymentStatusDuration
# * EstimatedReturn
# ListingCategory
# * LoanStatus
cor(select(loans,
           LoanOriginalAmount,
           Term,
           MonthlyLoanPayment,
           StatedMonthlyIncome,
           DebtToIncomeRatio,
           CreditScoreRangeLower,
           CreditScoreRangeUpper,
           EmploymentStatusDuration,
           EstimatedReturn,
           BorrowerRate,
           BorrowerAPR),
    y = NULL,
    use = 'pairwise.complete.obs')
```

Original amount correlates with monthly payment, term, credit score range lower
and credit score range upper.

Credit score range lower and credit score range upper are almost identical.
Borrower rate and borrower APR also have a very strong realtionship.
This means I can omit credit score range upper and borrower APR.
I am going to use credit score range lower and borrower rate only.

I would like to see how original amount and monthly payment depends on other
variables, and I am also interested in the changes of different variables
over the time.

### Relationships of Loan Original Amount and Monthly Loan Payment

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(MonthlyLoanPayment, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1)
```

There are three strong linear relationships between original amount and monthly
payment.

This means that original amount cannot describe the variation of monthly
payment alone, there must be something else participating in it.
Based on the strong correlation between term and monthly payment, and knowing
that there are three possible term values, term seems the best next variable
to investigate.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(MonthlyLoanPayment, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1, aes(color = factor(Term))) +
  scale_color_discrete() +
  legend_alpha_1
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(MonthlyLoanPayment, LoanOriginalAmount / Term), data = loans) +
  geom_jitter(alpha = .1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
with(loans, cor.test(MonthlyLoanPayment, LoanOriginalAmount / Term))
```

Monthly loan payment almost equals to the original amount divided by
the term of the loan.
Therefore monthly loan payment can be described by original
amount and term together.

Let's see how credit score and borrower rate values affect original amount and
monthly payment.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(CreditScoreRangeLower, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(CreditScoreRangeLower, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .1)
```

Regardless the relatively strong correlations I cannot find any visible
relationship between credit score range lower / borrower rate and
original amount / monthly payment.

The last plot seems overplotted, it may worth to revisit it with a lover
alpha level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .01)
```

With the decreased alpha level, I can see some underlying trend, but it is still
not enough clean.

Adding term to the plot may help again.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .01, aes(color = factor(Term))) +
  scale_color_discrete()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .01) +
  facet_wrap(~Term)
```

Faceting by term shows a some difference between the previously seen trends.

As term and original amount has a relationship to monthly loan payment
adding original amount to the previous plot may make the picture clearer.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .05, aes(color = LoanOriginalAmount)) +
  scale_color_continuous(low = 'light blue', high = 'dark blue') +
  facet_wrap(~Term)
```

Faceting by term and colouring by original amount shows a complex relationship,
were both borrower rate, orginal amount and term participates in
monthly payment.

I am going to check if stated monthly income can add anything to the picture.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(StatedMonthlyIncome, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1) +
  xlim(0, quantile(loans$StatedMonthlyIncome, .99))
```

It is clearly visible on this plot (as well), that there are some original
amount values that more frequent than the others (approximately every
five thousandth), but other than that, this plot cannot help too much.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(ListingCategory, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .1) +
  rotate_x
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(ListingCategory, MonthlyLoanPayment), data = loans) +
  geom_boxplot(aes(fill = ListingCategory)) +
  scale_color_discrete() +
  rotate_x
```

There is a relationship between monthly loan payment and listing category.
Loans for `Debt Consolidation` have the highest monthly payment, followed by
`Not Available`, `Home Improvement`, `Business` and `Other`.

#### Changes in Variables Over Time

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanOriginationDate, Term), data = loans) +
  geom_jitter(alpha = .05) +
  scale_y_continuous(breaks = c(12, 36, 60))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginationDate, color = factor(Term)), data = loans) +
  geom_freqpoly(size = 1, binwidth = 30) +
  scale_color_discrete() +
  scale_x_date()
```

All the loans before the previously mentioned gap (between late 2008 and late
2009) have a 36-months term.
12-months loans were only occurring for a 2-years period between 2011 and 2013
and were not popular.
60-months loans started to occur at the same time as 12-months loans (in 2011),
and there is an increasing trend in the number of those since then.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanOriginationDate, EmploymentStatus), data = loans) +
  geom_jitter(alpha = .05)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginationDate, color = EmploymentStatus), data = loans) +
  geom_freqpoly(size = 1, binwidth = 30) +
  scale_color_discrete() +
  scale_x_date()
```

There is no useful employment status information before late 2006.
The `Employed` category almost replaced the `Full Time` category in late
2010, this is probably due to a change in the way the data was recorded.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanOriginationDate, EmploymentStatusDuration), data = loans) +
  geom_jitter(alpha = .05, color = 'blue') +
  geom_smooth(color = 'black', size = 1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanOriginationDate, EmploymentStatusDuration), data = loans) +
  geom_smooth(aes(color = EmploymentStatus))
```

The employment status durations have an increasing trend, but it does
not have a strong relation with the changes int the employment status.

Let's see if there is any relationship between employment status and employment
status duration.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(EmploymentStatus, EmploymentStatusDuration), data = loans) +
  geom_boxplot(aes(fill = EmploymentStatus)) +
  scale_color_discrete() +
  rotate_x
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(EmploymentStatus, EmploymentStatusDuration), data = loans) +
  geom_jitter(alpha = .05) +
  rotate_x
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(EmploymentStatusDuration), data = loans) +
  geom_density(aes(color = EmploymentStatus), size = 1) +
  scale_color_discrete()
```

There is a relationship between the employment status and the employment status
duration.
`Employed`, `Full-time`, `Retired` and `Self-employed` statuses tend to have a
higher duration as `Not employed` or `Part-time`.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(LoanOriginationDate, ListingCategory), data = loans) +
  geom_jitter(alpha = .05)
```

There is no useful listing category information before 2008.
New listing categories were introduced in 2008 and 2012 as well.
This indicates a change in the way the data was recorded, similarly to listing
category.

There is no visible pattern within listing categories.

## Final Plots

### First Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(MonthlyLoanPayment, LoanOriginalAmount), data = loans) +
  geom_jitter(alpha = .1, aes(color = factor(Term))) +
  scale_color_discrete(name = 'Term') +
  legend_alpha_1 +
  ggtitle('Monthly Loan Payment by Loan Original Amount by and Term') +
  xlab('Monthly Loan Payment') +
  ylab('Loan Original Amount')
```

There is a very strong relationship between the term, the original amount and
the monthly payment of a loan.
The latter can be calculated from the formers.
The monthly payment of a loan increases as the term decreases - short term loans
have higher monthly payment than long term loans - and increases as
the original amount increases.

### Second Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(BorrowerRate, MonthlyLoanPayment), data = loans) +
  geom_jitter(alpha = .05, aes(color = LoanOriginalAmount)) +
  scale_color_continuous(low = 'light blue',
                         high = 'dark blue',
                         name = 'Loan Original Amount') +
  facet_wrap(~Term) +
  ggtitle(
    'Monthly Loan Payment by Borrower Rate, Term \n and Loan Original Amount') +
  xlab('Borrower Rate') +
  ylab('Monthly Loan Payment')
```

Monthly loan payment is affected by the borrower rate, higher borrower rate
leads to higher monthly payemnt.
This relationship is not affected by either the term or the original amount of
a loan.

### Third Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(ListingCategory, MonthlyLoanPayment), data = loans) +
  geom_boxplot(aes(fill = ListingCategory)) +
  ylim(0, 400) +
  scale_color_discrete() +
  guides(fill = FALSE) +
  rotate_x +
  ggtitle('Monthly Loan Payment by Listing Category') +
  xlab('Listing Category') +
  ylab('Monthly Loan Payment')
```

Monthly loan payment vary differently across listing categories.
Some categories has more wide monthly payment range (e.g., `Debt
Consolidation`), others has less wide payment range (e.g., `Motorcycle`).

## Reflection

The prosper loan dataset contains 113937 observations of 83 variables from
2006 to 2014. At the beginning of my investigation I was trying to understand
the meaning of different variables, find issues with how they stored and how
can I make the best use of them. As the dataset contains a lot of variables,
it was challenging to decide where to start. Finally, I have chosen to start
investigation the variables that best representing a loan: original amount,
term and monthly payment, and everything else that can have a relationship
with them. First I started to plot each variable one by one to have an
initial view about what the dataset contains, and how good the quality of it.
I was very surprised when I saw that there were loans with 0 monthly payment,
but later I realised that those loans are not current anymore. It was also
unexpected that there were no loans in the dataset between late 2008 and late
2009 but after some research I understood that the reason for the gap is the
[subprime mortgage crisis][cnn_subprime_mortgage_crisis]. Later, during the
analysis of the relationships between the different variables I realized that
the monthly payment is the most important property of a loan, instead of the
original amount as I thought before. I found that the term and the original
amount of a loan the most relevant parameters to calculate the monthly
payment. Borrower rate has an important role as well, however it was quite
hard to visualize that relationship. I also found that there is a difference
in the monthly payment, and the variation of the monthly payment across
listing categories. Analyzing the changes of the variables over the time
shown me that there were a few changes in the way how was the data collected.
That made it hard to use some categorical variables, like employment status.
The standardization of the mentioned categorical variables would lead to an
investigation that could make a better use of them. Investigating the dataset
without the loans that are for debt consolidation could give a good picture
of the choices of the borrowers. Investigating the loans that are paid by
debt consolidation would be useful as well.

[cnn_subprime_mortgage_crisis]: http://money.cnn.com/2008/09/15/news/economy/subprime_timeline/index.htm
